FATE summary results

### 2019-11-22 paper draft.Rmd
Tweaks to plankton processing

Now the zoops model.

```{r, echo=F}
GOA.zoop <- readRDS("/Users/MikeLitzow/Documents/R/Fate/fate-ewi/GOA_zoop_7.19.18.one.trend.rds") 
GOA.zoop$summary
```

@@

# load updated ewidata!
ewidata <- as.tibble(read.csv("ewidata.csv", row.names=1))

# drop salmon
drop <- grep("PI", GOA$code)
GOA <- GOA[-drop,]

drop <- grep("CO", GOA$code)
GOA <- GOA[-drop,]

# drop climate
drop <- grep("AKCLIM", GOA$code)
GOA <- GOA[-drop,]

unique(GOA$code)

# drop Pavlof and Chiniak and herring and kittiwakes
drop <- grep("PAV", GOA$code)
GOA <- GOA[-drop,]

drop <- grep("CHI", GOA$code)
GOA <- GOA[-drop,]

drop <- grep("HER", GOA$code)
GOA <- GOA[-drop,]

drop <- grep("BLKI", GOA$code)
GOA <- GOA[-drop,]

# and drop ichthyoplankton
drop <- grep("ICH", GOA$code)
GOA <- GOA[-drop,]

unique(GOA$code)
# recover the years and names for plots!
sub_data = GOA

@@

rotated = rotate_trends(GOA.zoop$best_model)

new.names <- c("CPR.copepods", "CPR.diatoms", "Icy.Str.amphipods", "Icy.Str.euphausids", "Icy.Str.gastropods", "CPR.mesozooplankton", "Seward.A.calanoid", "Seward.A.euphausid", "Seward.B.calanoid", "Seward.B.euphausid", "Seward.LA.euphausid", "Seward.LB.calanoid", "Seward.LB.euphausid", "Seward.calanoid.size", "Seward.euphausid.size", "Seward.LA.calanoid")

@@

plot.d <- ggplot(plot.load, aes(key, value)) +
  theme_bw() +
  geom_violin(fill=cb[3]) +
  geom_hline(yintercept = 0) +
  theme(axis.title.y = element_blank()) + ylab("Loading") +
  coord_flip() +
  ggtitle("d) Zooplankton loadings") +
  theme(title = element_text(size=8))

@@

plot.e <- ggplot(plot.trend, aes(year, trend)) +
  theme_bw() +
  geom_line(color=cb[3]) +
  geom_ribbon(aes(ymin=lo, ymax=hi), fill=cb[3], alpha=0.4) +
  geom_hline(yintercept = 0) + ylab("Trend value") + xlab("Year") +
  ggtitle("e) Zooplankton trend") +
  theme(title = element_text(size=8))

@@

zoop.regimes <- data.frame(year=plot.trend$year, probability=kk$data$median)

zoop.regimes$data.set <- "Zooplankton 1997-2016"

@@

years <- as.numeric(colnames(GOA.zoop$best_model$data))

@@

plot.f <- ggplot(mod, aes(x=SST)) +
  theme_bw() +
  geom_density(fill=cb[3]) + 
  geom_hline(yintercept=0) + 
  geom_vline(xintercept = 0, lty=2) + 
  xlab("Slope") + 
  ylab("Density") +
  ggtitle(("f) Zooplankton slope on SST")) +
  theme(title = element_text(size=8))

@@

```{r}
png("combined short time series plots.png", 8,5, units="in", res=300)

ggpubr::ggarrange(plot.a, plot.b, plot.c, plot.d, plot.e, plot.f,
          ncol=3, widths=c(1, 1, 0.8), nrow=2)

@@

>>>>

Now the zoops model.
These are now fit as three models- one for Seward Line, one for Icy Strait, one for large-scale (CPR & PP)
```{r, echo=F}
seward <- readRDS("Sewardline_11.19.19.one.trend.rds") 
seward$summary
```

@@

# load updated ewidata!
ewidata <- as.tibble(read.csv("ewidata.plankton.updated.csv"))

keep <- grep("SEWARD", ewidata$code)
seward.dat <- ewidata[keep,]
unique(seward.dat$code)

# recover the years and names for plots!
sub_data = seward.dat

@@

rotated = rotate_trends(seward$best_model)

new.names <- c("Calanoid.size.Sept", "Calanoid.size.May", "Euphausid.size.Sept", "Euphausid.size.May",
               "Calanoid.biomass.Sept", "Calanoid.biomass.May", "Euphausid.biomass.Sept", "Euphausid.biomass.May")

@@

plot.d <- ggplot(plot.load, aes(key, value)) +
  theme_bw() +
  geom_violin(fill=cb[3]) +
  geom_hline(yintercept = 0) +
  theme(axis.title.y = element_blank()) + ylab("Loading") +
  coord_flip() +
  ggtitle("d) Sewardline loadings") +
  theme(title = element_text(size=8))

@@

plot.e <- ggplot(plot.trend, aes(year, trend)) +
  theme_bw() +
  geom_line(color=cb[3]) +
  geom_ribbon(aes(ymin=lo, ymax=hi), fill=cb[3], alpha=0.4) +
  geom_hline(yintercept = 0) + ylab("Trend value") + xlab("Year") +
  ggtitle("e) Sewardline trend") +
  theme(title = element_text(size=8))

@@

seward.regimes <- data.frame(year=plot.trend$year, probability=kk$data$median)

seward.regimes$data.set <- "Sewardline 1998-2018"

@@

years <- as.numeric(colnames(seward$best_model$data))

@@

plot.f <- ggplot(mod, aes(x=SST)) +
  theme_bw() +
  geom_density(fill=cb[3]) + 
  geom_hline(yintercept=0) + 
  geom_vline(xintercept = 0, lty=2) + 
  xlab("Slope") + 
  ylab("Density") +
  ggtitle(("f) Sewardline slope on SST")) +
  theme(title = element_text(size=8))

Now the icy strait model.

```{r, echo=F}
icy <- readRDS("ICY_11.20.19.one.trend.rds") 
icy$summary
```

# load updated ewidata!
ewidata <- as.tibble(read.csv("ewidata.plankton.updated.csv"))

keep <- grep("ICY", ewidata$code)
icy.dat <- ewidata[keep,]
unique(icy.dat$code)

# drop total density
drop <- grep("TOTDEN", icy.dat$code)
icy.dat <- icy.dat[-drop,]

# recover the years and names for plots!
sub_data = icy.dat

rotated = rotate_trends(icy$best_model)

new.names <- c("Amphiphod", "Large.copepod", "Small.copepod", "Euphausid",
               "Gastropod")

# plot
# AND THERE IS ONLY ONE TS WITH > 0.95 LOADINGS!
plot.load <- loadings.1 %>%
  gather()

plot.d <- ggplot(plot.load, aes(key, value)) +
  theme_bw() +
  geom_violin(fill=cb[3]) +
  geom_hline(yintercept = 0) +
  theme(axis.title.y = element_blank()) + ylab("Loading") +
  coord_flip() +
  ggtitle("d) Sewardline loadings") +
  theme(title = element_text(size=8))

Now the large-scale!

```{r, echo=F}
large <- readRDS("PP_CPR_11.20.19.one.trend.rds") 
large$summary
```

# load updated ewidata!
ewidata <- as.tibble(read.csv("ewidata.plankton.updated.csv"))

keep <- c(grep("PP", ewidata$code), grep("CPR", ewidata$code))
large.dat <- ewidata[keep,]
unique(large.dat$code)

# recover the years and names for plots!
sub_data = large.dat

rotated = rotate_trends(large$best_model)

new.names <- c("Copepod.size", "Diatom.abundance", "Large.copepod.abundance", "Mesozoop.abundance",
               "Small.copepod.abundance", "Bloom.amplitude", "Bloom.duration", "Bloom.start")

plot.g <- ggplot(plot.load, aes(key, value)) +
  theme_bw() +
  geom_violin(fill=cb[3]) +
  geom_hline(yintercept = 0) +
  theme(axis.title.y = element_blank()) + ylab("Loading") +
  coord_flip() +
  ggtitle("g) Large-scale loadings") +
  theme(title = element_text(size=8))

plot.g
```

plot.h <- ggplot(plot.trend, aes(year, trend)) +
  theme_bw() +
  geom_line(color=cb[3]) +
  geom_ribbon(aes(ymin=lo, ymax=hi), fill=cb[3], alpha=0.4) +
  geom_hline(yintercept = 0) + ylab("Trend value") + xlab("Year") +
  ggtitle("h) Large-scale trend") +
  theme(title = element_text(size=8))

plot.h
```

Now test for SST effect.
```{r, include=F}
#  load sst as the explanatory variable.
ewidata <- read.csv("updated goa climate data.csv", row.names=1)

years <- as.numeric(colnames(seward$best_model$data))

plot.i <- ggplot(mod, aes(x=SST)) +
  theme_bw() +
  geom_density(fill=cb[3]) + 
  geom_hline(yintercept=0) + 
  geom_vline(xintercept = 0, lty=2) + 
  xlab("Slope") + 
  ylab("Density") +
  ggtitle(("i) Large-scale slope on SST")) +
  theme(title = element_text(size=8))

Look at era-specific effects.
```{r, include=F}
# first, define the function
# increasing to 4000 iterations!
trend_lm <- function(rotated_modelfit,
  all_years, selected_years,
  climate_dat = NULL,
  trend_number = 1, samples = 50,
  ITER = 4000, CHAINS = 1, n_eff = 50, rhat = 1.1, show_plot = TRUE) { # increasing from 1000 to 4000 iterations to help plankton model converge

years <- as.numeric(colnames(large$best_model$data)) # these are the years for the biology model - 1972:2016

cli_dat$slope <- 1
cli_dat$slope[cli_dat$Year %in% 2014:2019] <- 2
cli_dat$slope <- as.factor(cli_dat$slope)

set.seed(99)
mod = trend_lm(rotated_modelfit = r, all_years = 1998:2019,
  selected_years = 1998:2019, climate_dat=cli_dat)

names(mod)[2:3] <- c("1998-2013", "2014-2019")
summary(mod)

for(i in 2:3){
x <- sort(mod[,i])
print(names(mod)[i])
print(paste("median: ", median(x)))
print(paste("LCI: ", x[(0.025*25000)]))
print(paste("UCI: ", x[(0.975*25000)]))
}

melt.mod <- melt(mod, measure.vars=c("1998-2013", "2014-2019"))
names(melt.mod)[3] <- "era"
melt.mod$covar <- "SST"

# make a SST-only plot
ggplot(melt.mod, aes(x=value, fill=era)) +
  theme_bw() + 
  geom_density(alpha=0.6) +
  geom_vline(xintercept = 0, lty=2) +
  scale_fill_manual(values = cb[c(6, 2)]) +
  theme(legend.title = element_blank(), legend.position = c(0.8, 0.8)) +
  xlab("Slope") #+
  #xlim(-5,12)

ggsave("large area era sst slopes.png", width=5, height=4, units="in")

@@

```{r}
png("combined updated plankton time series plots.png", 8,7.5, units="in", res=300)

ggpubr::ggarrange(plot.a, plot.b, plot.c, plot.d, plot.e, plot.f, plot.g, plot.h, plot.i,
          ncol=3, widths=c(1, 1, 0.8), nrow=3)

@@

Or, much better! I ran a single model for the entire plankton TS and this appears to work...

```{r, echo=F}
all <- readRDS("all_plankton_11.20.19.one.trend.rds") 
all$summary
```

# load updated ewidata!
ewidata <- as.tibble(read.csv("ewidata.plankton.updated.csv"))

GOA <- ewidata

# select ICY, PP, Seward, CPR, and ichthyo!
keep <- c(grep("ICY", GOA$code), grep("PP_", GOA$code), grep("CPR", GOA$code), grep("SEWAR", GOA$code), grep("ICH", GOA$code))
GOA <- GOA[keep,]

rotated = rotate_trends(all$best_model)

new.names <- c("CPR copepod size", "CPR diatom", "CPR large copepod", "CPR mesozooplankton", "CPR small copepod", "Ichthyo. - sand lance", 
               "Ichthyo. - ronquils", "Ichthyo. - pollock", "Ichthyo. - Pacific cod", "Ichthyo. - flathead sole", "Ichthyo. - rock sole", "Ichthyo. - rockfishes", "Icy Strait amphipod",
               "Icy Strait large copepod", "Icy Strait small copepod", "Icy Strait euphausid", "Icy Strait gastropod", "Bloom amplitude", "Bloom duration", "Bloom start", "Seward calanoid size Sept.",
               "Seward calanoid size May", "Seward euphausid size Sept.", "Seward euphausid size May", "Seward calanoid Sept.", "Seward calanoid May", "Seward euphausid Sept.", "Seward euphausid May")

# and check
check <- data.frame(names, new.names)
check

# plot

plot.load <- loadings.1 %>%
  gather()

plot.a <- ggplot(plot.load, aes(key, value)) +
  theme_bw() +
  geom_violin(fill=cb[3], color=NA) +
  geom_hline(yintercept = 0) +
  theme(axis.title.y = element_blank()) + ylab("Loading") +
  coord_flip() +
  ggtitle("a) Plankton loadings") +
  theme(title = element_text(size=8)) +
  ylim(-2.5,2.5)

plot.a
```

And trend.

plot.b <- ggplot(plot.trend, aes(year, trend)) +
  theme_bw() +
  geom_line(color=cb[3]) +
  geom_ribbon(aes(ymin=lo, ymax=hi), fill=cb[3], alpha=0.4) +
  geom_hline(yintercept = 0) + ylab("Trend value") + xlab("Year") +
  ggtitle("b) Plankton trend") +
  theme(title = element_text(size=8))

plot.b
```

Now the slope on SST.

years <- as.numeric(colnames(all$best_model$data))

plot.c <- ggplot(mod, aes(x=SST)) +
  theme_bw() +
  geom_density(fill=cb[3]) + 
  geom_hline(yintercept=0) + 
  geom_vline(xintercept = 0, lty=2) + 
  xlab("Slope") + 
  ylab("Density") +
  ggtitle(("c) Plankton slope on SST")) +
  theme(title = element_text(size=8))

plot.c
```

```{r}
png("combined all plankton time series plots.png", 8, 2.5, units="in", res=300)

ggpubr::ggarrange(plot.a, plot.b, plot.c,
          ncol=3, widths=c(1, 1, 0.8), nrow=1)

Bimodal - evidence of nonstationary slope?

years <- as.numeric(colnames(all$best_model$data)) # these are the years for the biology model - 1972:2016

cli_dat$slope <- 1
cli_dat$slope[cli_dat$Year %in% 1989:2013] <- 2
cli_dat$slope[cli_dat$Year %in% 2014:2019] <- 3
cli_dat$slope <- as.factor(cli_dat$slope)

set.seed(99)
mod = trend_lm(rotated_modelfit = r, all_years = years[years %in% 1981:2019],
  selected_years = years[years %in% 1981:2019], climate_dat=cli_dat)

names(mod)[2:4] <- c("1981-1988", "1989-2013", "2014-2017")
summary(mod)

melt.mod <- melt(mod, measure.vars=c("1981-1988", "1989-2013", "2014-2017"))
names(melt.mod)[3] <- "era"
melt.mod$covar <- "SST"

# and plot
# make a SST-only plot
ggplot(melt.mod, aes(x=value, fill=era)) +
  theme_bw() + 
  geom_density(alpha=0.6) +
  geom_vline(xintercept = 0, lty=2) +
  scale_fill_manual(values = cb[c(4, 6, 2)]) +
  theme(legend.title = element_blank(), legend.position = c(0.8, 0.8)) +
  xlab("Slope") +
  xlim(-8,7)

ggsave("all plankton era sst slopes.png", width=5, height=4, units="in")
```
